### 2016-07-01 image_setup.tex
New pieces of the structure

\label{sec:testbed}

% We proceed to define the testbed environment by each of its conforming
% parts. We later indicate a procedure to set up and configure the testbed
% for several \ac{Raspi}s. In this procedure, we summarize all the key details
% regarding on setting the system files, configuration packages, tools and connectivity
% for the \ac{Raspi}s in a centralized fashion.

%Afterwards, we indicate how to cross-compile C++ code and the Kodo
%network coding library in an easy way.
%Finally, we provide further information about Kodo itself in terms
%of testing, other platforms supported and source code documentation
%for further references.

%\subsection{System Overview and Design Criteria}

A sketch of the testbed is depicted in Fig.~\ref{fig:testbed_setup}.
The testbed consists of up to 100 \ac{Raspi}s of different models.
More specifically, in our design we consider: Raspberry Pi 1 model B rev.~2,
Raspberry Pi 2 model B V1.1 and Raspberry Pi 3 model B V1.2.
All \ac{Raspi}s are each equipped with a memory card, a wired and wireless
network interface and a power supply. All the \ac{Raspi} are connected to a common
\ac{LAN} that provides internal and external connectivity. Without loss of generality,
in our case they are connected to the university network using their wired Ethernet interface
that is named \texttt{eth0} according to the legacy naming convention of Ethernet
interfaces in Linux~\cite{PredictableNetworkInterfaceNames}. We consider the
university network since we aim the testbed to be used by students and professors
to perform measurements and experimentation of controlled and reproducible scenarios
as part of academic research. We intend to configure all \ac{Raspi}s identically,
while still allowing the end-users to alter how each \ac{Raspi} behaves and is configured.

\begin{figure}[ht!]
\centering
\includegraphics[width=0.45\textwidth]{images/testbed_setup3.eps}
\caption{Testbed setup}
\label{fig:testbed_setup}
\end{figure}

%This significantly eases the management of the \ac{Raspi}s
%as \ac{SSH} eliminates the need for monitors and keyboards for the \ac{Raspi}s.

In Fig.~\ref{fig:testbed_setup} and what follows, we will refer
as the testbed administrator to the person(s) in charge of setting up and configuring
the testbed with administrator privileges from the \ac{OS} point of view.
The setting and configuration procedure commands are performed the testbed administrator in a
regular \ac{PC} running a Linux distribution. Although in principle the administrator
Linux distribution is not a restriction, we present our procedure in a Debian-based Linux
distribution. Finally, our design considers to fetch the \ac{Raspi}s
root filesystem from a \ac{HTTP} server. This enables the testbed administrator to
make filesystem in one location. The \ac{Raspi}s will then get the changes after a reboot.
All \ac{Raspi}s will be configured to run a \ac{SSH} daemon for easy remote access
within the university network. We requested the university \ac{IT} department to
configure the university \ac{DHCP} server to assign each \ac{Raspi} a static
\ac{IP}. This eliminates the requirements of monitors and keyboards with the
\ac{Raspi}s for non-graphical applications.

Based on a regular Raspbian Linux image, we explain how
to customize it to include desired packages for performing routinary tasks and set it
in all the \ac{Raspi}s. Also, when working on a testbed with many devices, it may not always be
desired to make persistent changes to the devices. Thus, we later present how to utilize
stacked filesystems to enable both persistent and temporary
storage. This step of the procedure will be optional if the testbed
administrator decides to write the whole image into a memory card. \todot{Check the
veracity and correctness of the previous statement. Question: Shoudln't we use
an overlay FS regardless of how the memory card is configured?}.
Although being simpler than a stacked filesystem, writing the whole Linux
filesystem to several memory cards may complicate future
maintenance of the testbed since updates need to be performed in each \ac{Raspi}.
Even when tools such as Python Fabric may ease this task, the maintenance might still
be cumbersome in some cases. Therefore, we propose and encourage the alternative centralized
approach where the initial \ac{RAM} filesystem is customized in a way that
enables \ac{Raspi}s to fetch the its root filesystem from a \ac{HTTP} server as
mentioned previously. The following procedure explains how the testbed may be setup and configured.

%\textcolor{red}{The testbed should be available for students to perform simulations and
%measurements as part of their semester projects. We intend to configure
%all \ac{Raspi}s identically, while still allowing students to alter
%how each \ac{Raspi} behaves and is configured.
%\\\\
%When working on the testbed with this many devices, it may not always be
%desired to make persistent changes to devices. We present how the concept
%of stacked filesystems may be used to enable the means of both persistent
%and temporary storage.}

%The testbed should be available for students to perform simulations and
%measurements as part of their semester projects. The \ac{Raspi}s should
%be configured identically, but allows the students to alter the
%configurations and how each \ac{Raspi} behaves.
%
%When working on a testbed with many devices, it may not always desired
%to make persistent changes to devices. We present how the concept of stacked
%filesystems may be used to enable the means of both persistent and
%temporary storage.

%A \ac{HTTP} server will be used to store Linux configuration files as well
%as the root filesystem in the centralized approach.

%The following procedure will explain how the testbed may be setup and
%configured both in the traditional distributed fasion where the
%entire Linux filesystem is written to an memory card, but also in a more
%centralized fasion where 
%is written to
%Based on the popular Debian based Raspbian Linux distribution
%We
%first present how a standard Linux distribution for the \ac{Raspi}s
%may be customized

\subsubsection{Installing Raspbian}

@@

\subsection{Configuring OS files and scripts}

The \ac{Raspi}s should be setup as similar as possible. One of those
few settings that may be different in the devices is their hostname.
%The hostname helps the user to distinguish the devices from each other.
%The hostname helps distinguish them from each other.
One approach to create unique hostnames would be to write the hostname
in each memory card. Another way is to create a start up script to automatically
configure the hostname. We decided to define the hostnames based on the \ac{MAC}
addresses of the \ac{Raspi}s wired ethernet interface. We store the \ac{MAC}
addresses and hostnames of the \ac{Raspi}s in a file as below. The testbed
administrator will need to insert the \ac{MAC} addresses and hostnames of
his/her \ac{Raspi}s.

%The \ac{Raspi}s should be setup to be as similar as possible. One of those
%things that is preferred to be different on the devices is their hostname.
%This helps distinguish them from each other. One solution to this in our
%approach would be to uniquely change the hostname in each memory card.
%This is however not desired because that would either require modifying
%the customized Raspbian lite image before writing the content to the
%memory card or to manually change it in the \ac{Raspi}s after the image has
%been written to a memory card.
%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.
%
%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames. This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.

%Below is a snippet of the file of all the hostnames and \ac{MAC} addresses:

%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames.
%This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp00
b8:27:eb:7b:c3:91 rasp01
b8:27:eb:54:9c:64 rasp02
b8:27:eb:95:bd:11 rasp03
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Provided that each \ac{Raspi} will have knowledge of all other \ac{Raspi}s in
%the testbed, it still needs to know which name to give itself.
%We do this using a small Bash script:
%We use the \ac{Bash} script below to enable each \ac{Raspi} to find its
%hostname in \texttt{nodes.csv}.
A \ac{Bash} script can then be used to have each \ac{Raspi} assign its hostname:

% Set hostname
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/set\_hostname}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

script_path="$(dirname $(realpath $0))"
config_file=${script_path}/nodes.csv 
mac=$(cat /sys/class/net/eth0/address)
old_hostname=$(hostname)
new_hostname=$(grep $mac $config_file | cut -f2 -d' ')

# Assign hostname found in nodes.csv
if [ ! -z ${new_hostname} ]; then
    echo ${new_hostname} > /etc/hostname
    hostname ${new_hostname}
    sed -i.old -e "s:${old_hostname}:${new_hostname}:g" /etc/hosts
fi
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The code lines above do the following:
\begin{itemize}
    \item Line 1: Tells system to interpret the script using \ac{Bash}
    \item Line 3-4: Gets the path to the script itself and the list of hostnames
    \item Line 5: Gets the \ac{MAC} address of the node itself
    \item Line 6: Gets the current hostname
    \item Line 7: Gets the new hostname from the hostname list
    \item Line 10-14: Assigns the new hostname
\end{itemize}

%Line 1 tells the system to interpret the script using Bash. Line 3 gets the
%path of the script itself. Line 4 gets the MAC address. This script will be
%located in all \ac{Raspi}s and thus return the MAC address of the node itself
%when executed.
%Line 4 searches line by line for this mac address.
%Line 8-11 assigns the hostname found in \texttt{"nodes.csv"}. If the MAC did not exists,
%then it will keep the default hostname of the system.

It is likely that the testbed administrator needs to either change or add
scripts to configure the \ac{Raspi}s. This would be tedious to distribute to
all \ac{Raspi}s at a later stage. We ease this task by making the \ac{Raspi}s fetch the
configuration scripts during start up before running the configuration scripts.

The below script attempts to download/update the scripts:

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/update\_rasp\_config}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

url="http://kom.aau.dk/project/TuneSCode/raspi/"
config_file="rasp_config.zip"

# Attempt to fetch new configuration files
if ! wget -q --show-progress -O /tmp/${config_file} ${url%/}/${config_file}; then
    echo "Warning: Unable to update rasp_config files"
    exit 1
fi

# Unzip and overwrite configurationn files to root's home directory
unzip -q -o /tmp/${config_file} -d /home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\begin{itemize}
    \item Line 3-4: Is the url and zip file that should be downloaded
    \item Line 7-10: Downloads the new configuration files to \texttt{/tmp}
        and prints a warning in case of errors
    \item Line 13: Unzips the files to \texttt{/home/pi/} while ignoring existing files/directories
\end{itemize}

We have packed the files and uploaded them to a webserver. These can be
retrieved with \texttt{wget} and unzipped into the Raspbian lite image:

% Get the files
\begin{lstlisting}[]
$ wget ${URL%/*}/rasp_config.zip
$ unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For the above scripts to work in the \ac{Raspi}s, it is required that the \ac{Raspi}
\ac{MAC} address is found in \texttt{nodes.csv}. Thus,
\texttt{rasp\_config.csv} needs to be unzipped, altered, and zipped again in
the \ac{HTTP} server from which the \ac{Raspi} retrieves the files from.
The testbed administrator will therefore need to also edit the \ac{URL} in the
script \texttt{update\_rasp\_config}.
%"nodes.csv" can be changed using a text editor. For vim, this could be done
%the following way:
%\begin{lstlisting}[]
%$ vim ${ROOTDIR}/home/pi/rasp_config/nodes.csv
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

Finally, to actually make the \ac{Raspi}s change hostname, we have to make
each \ac{Raspi} call the above script when it starts up. Optionally,
also run the update script before running other scripts.
Do this with root permissions by inserting the line(s) below before
\texttt{"exit 0"} using your favorite editor:
%"\$\{ROOTDIR\}/etc/rc.local" using
%your favorite text editor..
%The file
%can be opened with your favorite text editor.

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/set_hostname
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Configuring \texttt{rc.local} makes it really cumbersome for the testbed
administrator to add new scripts since it first needs to be included
in \texttt{rasp\_config.zip} and then each \ac{Raspi} needs to have an additional 
line added to \texttt{rc.local} in the root image or in all memory cards depending on
the final testbed setup.

A solution to this is to make yet another script in \texttt{rasp\_config.zip}
that calls all the other scripts in the desired order. This would look like
below:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/main}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

bash /home/pi/rasp_config/set_hostname
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to use the above script, \texttt{rc.local} needs to look like this:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/main
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
Notice that \texttt{set\_hostname} is now called by the \texttt{main}
script instead. The update script is still called directly. This ensures that
all configuration scripts are updated before executed. Changes to the update
script itself will first take effect at next system startup.
%Had the update script
%been called in \texttt{main} instead would mean that changes to the 
%\texttt{main} script would also first have been available at next startup.

%This can be done by inserting a call to the script
%set\_hostname in rc.local before it exits.
%
%Call set\_hostname script at startup. We insert a line of code to call script in rc.local just after the initial comments (i.e. lines starting with \#).
%\begin{lstlisting}[]
%# line_number=$(egrep -n -m1 "(^[^#])|(^$)" ${ROOTDIR}/etc/rc.local | cut -d: -f1)
%# sed -i "${line_number}c bash /home/pi/rasp_config/set_hostname" ${ROOTDIR}/etc/rc.local
%\end{lstlisting}
%\FloatBarrier
%$ line="bash /home/pi/.rasp_config/set_hostname"
%# awk -v text="$line" '!/^#/ && !p {print text; p=1} 1' ${ROOTDIR}/etc/rc.local > <@ \Suppressnumber @>
%    ${ROOTDIR}/etc/rc.local <@ \Reactivatenumber @>

%WE NEED TO ADD A FEW COMMANDS TELLING HOW TO ZIP, MODIFY, AND UNZIP
%FOR THE READER TO BE ABLE TO UPLOAD A CONSTRUCT HIS/HER OWN BUNDLE.

%All other files could be updated as well.
It is not only possible to add and configure files to/within the image. One can
also install and update the packages within the image which will be illustrated
in the next section.

\subsection{Installing and updating the image}

It may be desired to pre-install some programs in the image before it is
written to all the memory cards that goes into the \ac{Raspi}s.
This can be done from any linux x86 machine using QEMU Chroot~\cite{QemuUserEmulation}
(change root).
%\url{https://wiki.debian.org/QemuUserEmulation}
%\url{https://wiki.archlinux.org/index.php/Raspberry_Pi}

Due to the \ac{ARM} processor that \ac{Raspi}s are using, it is required to
install some additional software:
%Because \ac{Raspi}s are equipped with an \ac{ARM} processor 
%%it is not as
%straightforward as performing the task for two Linux systems of the same
%architecture.

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --display qemu-arm
qemu-arm (enabled):
     package = qemu-user-static
        type = magic
      offset = 0
       magic = \x7fELF\x01\x01\x01\x00\x00\x00\x00\x00\x00\x00\x00\x00\x02\x00\x28\x00
        mask = \xff\xff\xff\xff\xff\xff\xff\x00\xff\xff\xff\xff\xff\xff\xff\xff\xfe\xff\xff\xff
 interpreter = /usr/bin/qemu-arm-static
    detector = 
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
% This was neccessary on arch
%# update-binfmts --importdir /var/lib/binfmts/ --import

Make sure the second command writes "enabled" as the output above. If that 
is not the case, then try enabling it:

\begin{lstlisting}[]
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Provided that qemu-arm is enabled, we should now be able to change root (chroot)
into our Raspbian lite image. Change root is a method in Linux that enables
the root (/) to be changed. Thus, enabling a Linux installation within a
Linux installation.

There are a few commands to be performed before actually changing root into the
root partion of the image.
First, to get internet access from within the Raspbian lite image it is needed
to copy our \texttt{resolv.conf} into the filesystem. Line 1 change directory
into root filesystem of the image. The second line copies \texttt{resolv.conf}
from the testbed administrators \ac{PC} to the image.

% CHROOT to OS image
\begin{lstlisting}[]
$ cd $ROOTDIR
# cp /etc/resolv.conf ${ROOTDIR}/etc/resolv.conf
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, because of the \ac{ARM} architecture,
\texttt{/usr/bin/qemu-arm-static} needs to be copied into the image before continuing:

% CHROOT to OS image
\begin{lstlisting}[]
# cp /usr/bin/qemu-arm-static ${ROOTDIR}/usr/bin
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The final preparation before changing root is to populate the directories \texttt{proc},
\texttt{sys}, and \texttt{dev}:

% CHROOT to OS image
\begin{lstlisting}[]
# mount  -t proc proc proc/
# mount --bind /sys sys/
# mount --bind /dev dev/
# mount --bind /dev/pts dev/pts
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Finally, it is time to change root.
%Finally, the filesystem is ready for us to change root.
This can be done with \texttt{proot}. It may be required to install \texttt{proot}
using "\texttt{apt-get install proot}":

% CHROOT to OS image
\begin{lstlisting}[]
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Most available material online uses the more known \texttt{chroot} command as
written in the code block below instead of \texttt{proot}. This did not work
correctly in our machines, but we present the command as an alternative to
\texttt{proot} in case the testbed administrator should be in the opposite
situration where \texttt{proot} is not working.

% CHROOT to OS image
\begin{lstlisting}[]
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash    (ALTERNATIVE TO proot)
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If everything went well, we should now be \texttt{chrooted} into the Raspbian lite
filesystem under its root. Optionally, change the prompt title to indicate that it
is a \texttt{chrooted} environment:

% Optionally, we may create a unique prompt to indicate we have changed root
\begin{lstlisting}[]
# export PS1="(chroot) $PS1"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The Raspbian lite system image should now be possible to use almost as if it
had been booted in a \ac{Raspi}.
A major difference is that the testbed administrator's \ac{PC} is likely
significantly faster than a \ac{Raspi}.
%The only difference is that the laptop that we are using is
%significantly faster than a \ac{Raspi}.
Thus enabling updates, upgrades, and installing new software packages much
faster than in the \ac{Raspi}s.
%Thus enabling us to for example update,
%upgrade, and install software packages:

% Update system
\begin{lstlisting}[]
(chroot) # apt-get update
(chroot) # apt-get upgrade
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

We also install a few packages that we find useful:
%Lets install some useful applications:
% Install packages
\begin{lstlisting}[]
(chroot) # apt-get install vim git screen
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

When writing the image to a memory card, all these change that has been made
to the image so far will exists in all \ac{Raspi}s.
%All the changes that are made here will exists in all \ac{Raspi}s when the
%image is written to a memory card.

\subsection{Installation an overlaying filesystem}

Our goal was to enable the users of the testbed to change however the
\ac{Raspi} are configured while still being able to revert those changes easily.

We do this by storing a users work in the \ac{Raspi}s in \ac{RAM}. Thus a
\ac{Raspi}, when a \ac{Raspi} is rebooted, it will forget all changes.

A way to do this is using stacked filesystem. Stacking filesystem enables
a filesystem to overlay another filesystem. Whenever a file is accessed,
the upper filesystem will forward the request to the lower filesystem in
case it does not have it itself. If it has it, it will simpy return the file.

This idea can be used in our setup to mount the root filesystem (i.e. Raspbian
lite) in the \ac{Raspi}s duing startup as read-only. Thus, the files will not
be changable. Some programs needs to change files and the testbed admin might
want to change configurations. To enable this, it is possible to
overlay a filesystem that is mounted in \ac{RAM} as read/writable on top of
the regular root. Thus, reading a file may return a file from the lower filesytem,
but if it is stored, it will be saved in the upper filesystem. Accessing this file
again will return the stored file from the upper layer.

%We do this using an overlay filesystem. The idea is to let the \ac{Raspi}s
%mount their root filesystem as 
%
%We do this using an overlay filesystem (overlayfs).
%Overlayfs enables a filesystem to be stacked on top of another filesystem. Thus,
%if the \ac{Raspi}s mount their root filesystem (/) as read-onlyif we mount our root (/) filesystem as 
%
%
%Overlayfs enables filesystems
%to be stacked on top of each other. We will be using this to make a lower
%filesystem and an upper filesystem. The lower filesystem will be our
%traditional filesystem and the upper filesystem will be mounted in tmpfs
%(i.e. mounted in \ac{RAM}).
%
%Thus, when a file is requsted from stacked filesystem, it will first attempt
%to fetch the file in the upper filesystem if it does not exists there, it will
%request it from the lower filessytem.

Assuming that you are still \texttt{chrooted} into the Raspbian lite filesystem, we
can setup the overlay filesystem (overlayfs).

There already exists implementations overlaying the root filesystem. We will
use an existing implementation
available at github: \cite{overlayroot}.

We installed git in a previous step, so lets clone the repository. The below
code stores it in /tmp which is really mounted in \ac{RAM}. Thus, the files
will disappear when the machine is rebooted.

% Get files
\begin{lstlisting}[]
(chroot) # OVERLAYROOTDIR="/tmp/overlayroot"
(chroot) # git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

To enable the overlaying filesystem it is required to generate an initial \ac{RAM}
file system (initramfs). Initramfs is an initial filesystem that gets loaded into
\ac{RAM} during the startup process of a linux machine to prepare the real
filesystem.

For this purpose, we need BuxyBox:

% Install required packages
\begin{lstlisting}[]
(chroot) # apt-get install busybox
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Add and activate the overlaying filesystem, we need to add the scripts and tell initramfs
to load the overlaying filesytem:

% Setup initramfs
\begin{lstlisting}[]
(chroot) # cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
(chroot) # cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
(chroot) # echo "overlay" > /etc/initramfs-tools/modules
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Finally, lets generate the initial \ac{RAM} filesystem. This is done using the command
%\texttt{mkinitramfs}, that will by default look for the kernel modules of the current system. Remember we are
%chrooted, so we need to specify the correct kernel modules to look for. These
%can be found in "/lib/modules".
Now, the initial \ac{RAM} filesystem can be generated. \ac{Raspi} version 1
needs a different kernel than \ac{Raspi} version 2 and version 3. Thus, start
by finding the available kernel modules:
%Because \texttt{initramfs} is executed within a
%\texttt{chrooted} environment, it will not be able to detect the
%kernel modules in the \texttt{chrooted} environment. The kernel modules will

% Create an initramfs
\begin{lstlisting}[]
(chroot) # ls /lib/modules/
4.4.13+  4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Kernel "4.4.13+" is for \ac{Raspi} version 1 and kernel "4.4.13-v7+" works
for \ac{Raspi} version 2 and version 3.

Lets proceed by generating an initial \ac{RAM} filesystem for both kernels:

%Lets proceed with kernel modules "4.4.13+".
%Lets create an initial ram filesystem for both kernels.

%Thus, to run mkinitramfs, type in the correct kernel version:

% Create an initramfs
\begin{lstlisting}[]
(chroot) # mkinitramfs -o /boot/init.gz -k 4.4.13+
(chroot) # mkinitramfs -o /boot/init-v7.gz -k 4.4.13-v7+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%IT SEEM THAT THE RASPIS DEMANDS DIFFERENT KERNEL MODULES AFTER ALL.
%RASP-VERSION1=4.4.13+, RASP-VERSION2=4.4.13-V7+, RASP-VERSION3=4.4.13-V7+

Although there might be printet a few warnings, the above commands should successfully
generate working initial \ac{RAM} filesystems.

An initial \ac{RAM} filesystem needs to be called by the bootloader. In Raspbian
this is done by adding a command to \texttt{config.txt} in the boot partition.

If the system should be run in a \ac{Raspi} version 1, then use \texttt{init.gz}:
%The bootloader should be made aware of the initramfs that we will generate:
\begin{lstlisting}[]
(chroot) # echo "initramfs init.gz" >> /boot/config.txt     # Only for Raspberry Pi version 1
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If the system should be run in a \ac{Raspi} version 2 or 3, then use \texttt{init-v7.gz}:
\begin{lstlisting}[]
(chroot) # echo "initramfs init-v7.gz" >> /boot/config.txt  # Only for Raspberry Pi version 2 and 3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

At this point, it is no longer required to be \texttt{chrooted} and the image
can be written to a memory card. The following commands exits the \texttt{chrooted}
environment, unmount all partitions, and detaches the loopback device:

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
(chroot) # exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{(Optional) Persistent subdirectories for distributed Linux image}

EMPHASIZE THAT THIS SETUP IS ONLY FOR THE SETUP WHERE THE ROOT IMAGE ARE
WRITTEN TO THE MEMORY CARDS

Provided the stacked filesystem was configured, it may be desired to have
some directories where files are not flushed upon rebooting the \ac{Raspi}s.
The following procedure will create an extra partition in the image for
the \ac{Raspi} users home directory that will be made storage persistent.
%Provided that a stacked filesystem was configured to make the root
%filesystem forget chances upon reboots is not always desired.

Expand image according to the desired home directory size. Keep in mind not
to make the image bigger than the target memory card.
% Expand image
\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create home partition after the root partition. Use \texttt{fdisk} to
find the next available sector in the image. In our case, the new
partition should start at sector 4806656:
\begin{lstlisting}[]
$ fdisk -lu ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 3.3 GiB, 3534749696 bytes, 6903808 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 4806655 4669440  2.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create new partition:

\begin{lstlisting}[]
$ (echo n; echo p; echo ; echo 4806656; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create loopback device: 
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Format new partition:
\begin{lstlisting}[]
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The last thing to remember is to inform Raspbian to mount the home partition
we created during startup. This can be done by adding an entry in \texttt{fstab}:
\begin{lstlisting}[]
# mount ${DEV}p2 ${ROOTDIR}
# sudo sh -c "echo '/dev/mmcblk0p3 /home ext4 defaults,noatime 0 2' >> ${ROOTDIR}/etc/fstab"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The home folder was originally located in the root filesystem. We should
therefore copy the content to the new home partition:
\begin{lstlisting}[]
# mount ${DEV}p3 ${ROOTDIR}/mnt
# mv ${ROOTDIR}/home/* ${ROOTDIR}/mnt/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

That is it. unmount partitions and detach loop device:
\begin{lstlisting}[]
# umount --recursive ${ROOTDIR}
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{Write image to memory card}

EMPHASIZE THAT THIS SETUP IS ONLY FOR THE SETUP WHERE THE ROOT IMAGE ARE
WRITTEN TO THE MEMORY CARDS

The final step is to write the image to all the memory cards before they can
be used in the \ac{Raspi}s.

%This is done by first leaving the \texttt{chrooted}
%filesystem and unmounting all the directories we mounted:

% Exit chroot, umount, and write to memory card
%\begin{lstlisting}[]
%# exit
%# cd ..
%# umount --recursive ${ROOTDIR}
%# losetup -d ${DEV}.img
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

%If successfully unmounted, you can 
Insert a memory card in your machine. It will be available as
\texttt{/dev/mmcblkX} or \texttt{/dev/sdX}

%In most systems, this memory card was available under \texttt{/dev/mmcblk0}, but this may be
%different on different machines. For internal memory card readers, it will
%probably be named /dev/mmcblkX where X is some number and if the memory card
%reader is connected through \ac{USB}, then it is likely called /dev/sdX.
Make absolutely sure to write to the correct device as everything will be 
overwritten!
Inserting/Ejecting the memory card a couple of times and using the commands
below may help deducing the correct device:
%The commands below may help deduce which device is the correct:
%Below is a few commands that may help you to deduce which device is correct:

\begin{lstlisting}[]
# lsblk
# df -h
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For us, the device was \texttt{/dev/mmcblk0}. Write the image to a memory card:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The memory card can now be started in a \ac{Raspi}.

>>>>

\label{sec:image_setup}

\subsection{Installing Raspbian}

@@
